Topic modeling on EU and the USA research projects

### 2017-06-12 thesis_report/draft_thesis.tex
small fixes in theoretic background & citations

\setlength{\parskip}{1em}

\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{algorithm2e}
\usepackage{url}

@@

In this chapter we will present some basic theoretic background about topic
modeling. To deal with extensive text body, machine learning researchers have
developed topic modeling: a suite of methods that analyze the words of the
original text to discover the themes that run through the data and detect
hidden semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can be used
for organization, finding similar documents, or any other similar tasks. A
topic is here defined as a probability distribution of words, meaning that
certain words are more likely within a certain topic. Since the topics emerge
from the analysis of the original texts, topic modeling algorithms do not
require any prior annotations or labeling of the documents.

@@

Some further common concepts and terms in topic modeling:

- A word is defined as an item from a vocabulary indexed from ${1, ..., V}$,
where V is the size of the vocabulary. All words are represented as unit-basis
vectors with one component equal to one and the rest equal to zero.

- A document is a collection of words denoted by $w = {w_1, ..., w_N}$, where
$w_n$ is the $n$th word and $N$ is the total number of words in the collection.

- A corpus is a collection of documents in a dataset. It is denoted $D = {w_1,
..., w_M}$, where $w_m$ is the $m$th document in the corpus and $M$ is the
total number of documents.

- Latent variables are variables that may not be directly observed, unlike
observable variables. Latent variables can instead be inferred from other
observable variables.

- Polysemy is the capacity for a word to have multiple related meanings. An
example of this is the word “plant” which can mean a living organism of the
kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.

- Synonymy is the capacity for several words to have similar meanings such as
the words “buy” and “purchase”.

@@

When introducing the Latent Dirichlet Allocation in section 2.x.x, we will be
using the terms prior and posterior probabilities and likelihood function:

@@

The probability distribution that we use is the multinomial. In probability 
theory, the multinomial distribution is a generalization of the binomial 
distribution. For example, it models the probability of counts for rolling 
a k-sided dice n times. For n independent trials each of which leads to a 
success for exactly one of k categories, with each category having a given 
fixed success probability, the multinomial distribution gives the 
probability of any particular combination of numbers of successes for the 
various categories \footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words ~\cite{Huang_maximumlikelihood}.

@@

The Dirichlet process is a way to model randomness of a probability mass
function (PMF) with unlimited options (e.g. an unlimited amount of dice in a
bag). The process is similar Polya’s Urn \footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}}, only instead of having a set number
of ball colors you have an unlimited amount:

@@

After LSI was introduced, Hofmann presented the Probabilistic Latent Semantic
Indexing, or PLSI model. PLSI is a so-called topic model where each word in a
document is generated from a single “topic” which results in that each document
in a corpus can be represented with a topic distribution. Later, in 2003 Blei et
al. presented the Latent Dirichlet Allocation, or LDA. As opposed to PLSI, LDA
is a statistically generative model for documents where each word in a document
can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in them. In LSI a
term-document matrix is created by analyzing the corpus, where the rows
correspond to words and columns to documents. Each element in this sparse
matrix describes the number of times a word occurs in a document but this term
count can also be weighted with for instance TF-IDF. Letting each word
represent a dimension in a very high dimensional space, a document can be seen
as a vector with components corresponding to its weighted term counts ~\cite{Deerwester90indexingby}.

@@

Probabilistic Latent Semantic Indexing evolved from LSI and uses the same
concept of finding a lower rank approximation of the term-document occurrence
matrix. The difference is that instead of being based on linear algebra, PLSI
is based on a mixture decomposition using a latent class model 
~\cite{Hofmann:1999:PLS:312624.312649}. PLSI associates
an unobserved class variable $z$ with each document-word observation pair $(w,
w)$. This $z$ can be seen as a topic, since it is a probability distribution
over words. As a generative model, PLSI can be defined in the following way for
a corpus $D$:

\begin{enumerate}
\item Pick a document $w$ with probability $p(w)$
\item For each of the $N$ words in $w$:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|w)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}

The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|w)$. Even though PLSI is generative for the
analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data. Another problem is that the number of parameters in the
model grows linearly with the number of documents.

@@

LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics. This is similar to PLSI, except
that in LDA the topic distribution is assumed to have a Dirichlet prior (see
2.3.3.1). In practice, this results in more reasonable mixtures of topics in a
document. The reduced document description result in PLSI is a vector where
each element describes the mixing proportion for a topic. A limitation with
PLSI is that there is no generative model for these proportions, making it
difficult to handle unseen documents. The Latent Dirichlet Allocation model
tries to solve this limitation by setting a Dirichlet prior on the topic
distribution ~\cite{blei2003latent}.

@@

The basic idea in LDA is that we define a topic to be a Dirichlet distribution
over a fixed vocabulary. Technically, the model assumes that the topics are
generated first, before the documents ~\cite{Blei11introductionto}. Now for each document in the collection,
we generate the words in a two-stage process:

\begin{enumerate}
\item Randomly choose a distribution over topics.
\item For each word in the document:
\begin{description}
	\item (a) Randomly choose a topic from the distribution over topics in step 1.
\end{description}
\begin{description}
	\item (b) Randomly choose a word from the corresponding distribution over the 
	vocabulary.
\end{description}
\end{enumerate}

This statistical model reflects the intuition that documents exhibit multiple
topics. Each document exhibits the topics with different proportion (step 1);
each word in each document is drawn from one of the topics (step 2b), where the
selected topic is chosen from the per-document distribution over topics (step
2a). All the documents in the corpus share the same set of topics, but each
document exhibits those topics with different proportion - that’s the
distinguishing characteristic of LDA. A more detailed explanation of the model
follows in section 2.3.3.1.

@@

In LDA, each document is represented by a mixture of the topics, with weight
$\theta_w^2$ for topic $z$ in document $w$. These weight have a Dirichlet prior,
parameterized by a hyper-parameter $\alpha$. Each topic is a probability
distribution over the vocabulary, with probability $\beta^w_z$ for word $w$ in
topic $z$.

For each document $w$, in corpus $D$, the LDA generative process is as follows:

\begin{enumerate}
	\item Draw topic distribution $\theta_w \sim Dir(\alpha)$
	\item For each of the $N$ words $w_n$:
	\begin{description}
	\item (a) Draw a specific topic $z_n \sim Multinomial(\theta)$
	\item (b) Draw a word $w_n \sim Multinomial(\beta_{zn})$
	\end{description}
\end{enumerate}

@@

A limitation with the standard LDA model is the fix vocabulary. This means that
new words cannot enter the model and that you have to know a priori what words
will model the corpus effectively. A setting where this could be problematic is
when analyzing twitter data where new words in the form of slang, names and
hashtags emerge over time. Another limitation is the fact that one has to state
the number of topics the model should use. This is not a big issue, but could
lead to some of the resulting topics to be insignificant.

\begin{center}
\begin{figure}
\centering
\includegraphics[width=0.6\textwidth]{LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously 
~\cite{Myung:2003}.

@@

Learning the various distributions (the set of topics, their associated word
probabilities, the topic of each word, and the particular topic mixture of each
document) is a problem of Bayesian inference. In the original paper it's being
used a variational Bayes approximation of the posterior distribution;
~\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed ~\cite{Fox2011ATO}.

@@

An online variational inference algorithm is proposed by Hoffman et all.
~\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The "online LDA" is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.

>>>>

\setlength{\parskip}{1em}
\usepackage[justification=centering]{caption}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{algorithm2e}
\usepackage{url}

@@

In this chapter we will present some basic theoretic background about topic
modeling. To deal with extensive text body, machine learning researchers have
developed topic modeling: a suite of methods that analyze the words of the
original text to discover the themes that run through the data and detect
hidden semantic structures by analyzing a corpus. By marking up the documents
using these topics and by using the resulting structure, the model can be used
for organization, finding similar documents, or any other similar tasks. A
topic is here defined as a probability distribution of words, meaning that
certain words are more likely within a certain topic. Since the topics emerge
from the analysis of the original texts, topic modeling algorithms do not
require any prior annotations or labeling of the documents.\cite{Blei:2012:PTM:2133806.2133826}

@@

Some further common concepts and terms in topic modeling: 
\begin{itemize}
\item[] A \textit{word} is defined as an item from a vocabulary indexed from 
${1, ..., V}$, where V is the size of the vocabulary. All words are 
represented as unit-basis vectors with one component equal to one and 
the rest equal to zero.
\item[] A \textit{document} is a collection of words denoted by $w = {w_1, ..., w_N}$, 
where $w_n$ is the $n$th word and $N$ is the total number of words in the 
collection.
\item[] A \textit{corpus} is a collection of documents in a dataset. It is denoted 
$D = {w_1,..., w_M}$, where $w_m$ is the $m$th document in the corpus and 
$M$ is the total number of documents.
\item[] \textit{Latent variables} are variables that may not be directly observed, 
unlike observable variables. Latent variables can instead be inferred 
from other observable variables.
\item[] \textit{Polysemy} is the capacity for a word to have multiple related meanings. 
An example of this is the word “plant” which can mean a living organism of 
the kind exemplified by trees, herbs etc., or a place where an industrial or
manufacturing process takes place.
\item[] \textit{Synonymy} is the capacity for several words to have similar meanings such as
the words “buy” and “purchase”.
\end{itemize}

@@

When introducing the Latent Dirichlet Allocation in section 2.2.3, we 
will be using the terms prior and posterior probabilities and 
likelihood function:

@@

The probability distribution that we use is the multinomial. In probability 
theory, the multinomial distribution is a generalization of the binomial 
distribution. For example, it models the probability of counts for rolling 
a k-sided dice n times. For n independent trials each of which leads to a 
success for exactly one of k categories, with each category having a given 
fixed success probability, the multinomial distribution gives the 
probability of any particular combination of numbers of successes for the 
various categories\footnote{\url{https://en.wikipedia.org/wiki/Multinomial_distribution}}.

@@

Like any discrete distribution, a multinomial has to sum to one, where the sum
is over all possible data points. Here, a data point is a document containing
$n$ words.\cite{Huang_maximumlikelihood}

@@

The Dirichlet process is a way to model randomness of a probability mass
function (PMF) with unlimited options (e.g. an unlimited amount of dice in a
bag). The process is similar Polya’s Urn\footnote{\url{http://www.statisticshowto.com/dirichlet-distribution/}}, only instead of having a set number
of ball colors you have an unlimited amount:

@@

After LSI was introduced, Hofmann presented the Probabilistic Latent Semantic
Indexing, or PLSI model. PLSI is a topic model where each word in a
document is generated from a single topic which results in that each document
in a corpus can be represented with a topic distribution. Later, in 2003 Blei et
al. presented the Latent Dirichlet Allocation, or LDA. As opposed to PLSI, LDA
is a statistically generative model for documents where each word in a document
can be generated by all topics.

@@

Latent Semantic Indexing was one of the earliest methods for finding
relationships between documents and the words that occur in them.\cite{Deerwester90indexingby} In LSI a
term-document matrix is created by analyzing the corpus, where the rows
correspond to words and columns to documents. Each element in this sparse
matrix describes the number of times a word occurs in a document but this term
count can also be weighted with for instance TF-IDF. Letting each word
represent a dimension in a very high dimensional space, a document can be seen
as a vector with components corresponding to its weighted term counts.
\cite{Salton:1988:TAA:54259.54260}

@@

Probabilistic Latent Semantic Indexing evolved from LSI and uses the same
concept of finding a lower rank approximation of the term-document occurrence
matrix. The difference is that instead of being based on linear algebra, PLSI
is based on a mixture decomposition using a latent class model.
~\cite{Hofmann:1999:PLS:312624.312649} PLSI associates an unobserved class 
variable $z$ with each document-word observation pair (\textbf{w},$
w)$. This $z$ can be seen as a topic, since it is a probability distribution
over words. As a generative model, PLSI can be defined in the following way 
for a corpus $D$:

\begin{enumerate}
\item Pick a document \textbf{w} with probability $p($\textbf{w}$)$
\item For each of the $N$ words in \textbf{w}:
\begin{description}
\item (a) Pick a topic $z$ with probability $p(z|$\textbf{w}$)$
\end{description}
\begin{description}
\item (b) Generate a word $w$ with probability $p(w|z)$
\end{description}
\end{enumerate}
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth, natwidth=548,natheight=201]{figs/PLSI_plate_notation.png}
\caption{Plate notation representing the PLSI model. \textbf{w} is the document index, $z$ is the topic drawn for the word $w$ from the topic distribution of the document}
\end{figure}
\end{center}
The result from PLSI is that every document is represented as mixing proportions
for the topics, given by $p(z|$\textbf{w}$)$. Even though PLSI is generative for 
the analyzed corpus, it is not generative for new documents, which means is that
there is no clear way of assigning probability to a document that is not part
of the training data.\cite{blei2003latent} Another problem is that the number 
of parameters in the model grows linearly with the number of documents.

@@

LDA is a topic model that was first presented as a graphical model for topic
discovery by David Blei, Andrew Ng, and Michael I. Jordan in 2003. The basic
idea is that documents exhibit multiple topics. LDA is a generative statistical
model that allows sets of observations to be explained by unobserved groups
that explain why some parts of the data are similar. For example, if
observations are words collected into documents, it posits that each document
is a mixture of a small number of topics and that each word's occurrence is
attributable to one of the document's topics.\cite{Blei:2012:PTM:2133806.2133826}

This is similar to PLSI, except
that in LDA the topic distribution is assumed to have a Dirichlet prior. 
In practice, this results in more reasonable mixtures of topics in a
document. The reduced document description result in PLSI is a vector where
each element describes the mixing proportion for a topic. A limitation with
PLSI is that there is no generative model for these proportions, making it
difficult to handle unseen documents. The Latent Dirichlet Allocation model
tries to solve this limitation by setting a Dirichlet prior on the topic
distribution.\cite{blei2003latent}

@@

The basic idea in LDA is that we define a topic to be a Dirichlet distribution
over a fixed vocabulary. Technically, the model assumes that the topics are
generated first, before the documents.\cite{Blei11introductionto} Now for 
each document in the collection, we generate the words in a two-stage process:

\begin{algorithm}[H]
\SetAlgoNoLine
1. Randomly choose a distribution over topics.

2. 
\For{each word in the document}{
	(a) Randomly choose a topic from the distribution over topics in 
	step 1

	(b) Randomly choose a word from the corresponding distribution over 
	the vocabulary
}
\end{algorithm}

This statistical model reflects the intuition that documents exhibit multiple
topics. Each document exhibits the topics with different proportion (step 1);
each word in each document is drawn from one of the topics (step 2b), where the
selected topic is chosen from the per-document distribution over topics (step
2a). All the documents in the corpus share the same set of topics, but each
document exhibits those topics with different proportion - that’s the
distinguishing characteristic of LDA. A more detailed explanation of the model
follows in section 2.2.3.1.

@@

In LDA, each document is represented by a mixture of the topics, with weight
$\theta_w^2$ for topic $z$ in document $w$. These weight have a Dirichlet prior,
parameterized by a hyper-parameter $\alpha$. Each topic is a probability
distribution over the vocabulary, with probability $\beta^w_z$ for word $w$ in
topic $z$. We define the following variables and notation:
\begin{itemize}
  \item[] $k$ is the number of topics.
  \item[] $V$ is the number of unique words in the vocabulary.
  \item[] $\theta$  is the topic distribution (of length $k$) for a document,
   drawn from a uniform Dirichlet distribution with parameter $\alpha$.
  \item[] $z_{n}$ is a topic assignment for $word w_{n}$, sampled from 
  $p(z_{n} = i|\theta) = \theta_{i}$.
  \item[] \textbf{w}$ = (w_{1}, ... , w_{N})$ is a document with $N$ words.
  \item[] $w_{n}^{i} = 1$ means that the word $w_{n}$ is the $i$-th word 
  of the vocabulary.
  \item[] $\beta$  is a $k \times V$ matrix, where each row $\beta_{i}$ 
  is the multinomial distribution for the $i$-th topic. That is, 
  $\beta_{ij} = p(w^{j} = 1 | z_{j} = i)$.
\end{itemize}

For each document $w$, in corpus $D$, the LDA generative process is as follows, 
with corresponding plate notation in Figure 2:

\begin{algorithm}[H]
\SetAlgoNoLine
1. Draw topic distribution $\theta_w \sim Dir(\alpha)$

2. 
\For{each of the $N$ words $w_n$}{
	(a) Draw a specific topic $z_n \sim Multinomial(\theta)$

	(b) Draw a word $w_n \sim Multinomial(\beta_{zn})$
}
\end{algorithm}

@@

A limitation with the standard LDA model is the fix vocabulary. This means that
new words cannot enter the model and that you have to know a priori what words
will model the corpus effectively. A setting where this could be problematic is
when analyzing data where new words in the form of slang or names emerge over 
time. Another limitation is the fact that one has to state
the number of topics the model should use. This is not a big issue, but could
lead to some of the resulting topics to be insignificant.
\begin{center}
\begin{figure}[h]
\centering
\includegraphics[width=0.6\textwidth, natwidth=561,natheight=197]{figs/LDA_standard_model.png}
\caption{Plate notation of the standard LDA model}
\end{figure}
\end{center}

@@

The Expectation Maximization, or EM, algorithm is an approach to deriving an
algorithm for approximately obtaining maximum likelihood of variables in a
statistical model. Normally one obtains a maximum likelihood solution by taking
partial derivatives of the likelihood function with respect to all variables,
setting these derivatives to zero and solving the equations simultaneously.
\cite{Myung:2003}

@@

Learning the various distributions (the set of topics, their associated word
probabilities, the topic of each word, and the particular topic mixture of each
document) is a problem of Bayesian inference. In the original paper it's being
used a variational Bayes approximation of the posterior distribution;\cite{blei2003latent} alternative inference techniques use Gibbs sampling and
expectation propagation. The goal of variational inference is to approximate an
intractable probability distribution $p$, with a tractable one $q$, in a way
that makes them as close as possible given a similarity measure. In this thesis,
variational inference is used to approximate the posterior distribution of
latent variables. A new distribution over the hidden variables $q(z, \beta)$
(called the variational distribution) is  introduced. This new distribution has
properties such that it can be efficiently computed.\cite{Fox2011ATO}

@@

An online variational inference algorithm is proposed by Hoffman et all.\cite{onlineLDAvb} for fitting $\lambda$, the parameters to the variational
posterior over the topic distributions $\beta$. The "online LDA" is nearly as
simple as the variational algorithm, but converges much faster for large
datasets. Variational inference requires to pass through the whole dataset for
each iteration, and that can be a problem when working with big datasets and is
not an option to use cases where the data is arriving constantly. The Online
Variational Inference algorithm aims to solve both problems without compromising
the quality of the obtained topics and with an algorithm nearly as simple as
Variational Inference. For this algorithm we will only highlight the differences
respect the Variational Inference and the resulting algorithm. The key
difference is the way to fit the variational parameter $\lambda$ respect the
Variational Inference explained in the previous section.
